---
title: "NYC Air Quality"
author: "Minh Tran"
date: "11-14-2024"
output: html_document

---
# New York City air quality surveillance report

2024-11-06

## Introduction

In our project, we analyze if air quality in New York City increased
during last years. For these purposes, we use New York City air quality
surveillance data (https://catalog.data.gov/dataset/air-quality). We 
investigate dynamics of 18 indexes of air quality. Some of them are considered
separately for adults and children. Our dataset consists of 18,025 observations.
One observation is value for some index for some period for some geo place.
Respectively, values for all indexes that are presented on plots are averaged
by geo places. In general, our data covers period 2005-2022 but for some
indexes it is narrower.

## Data preparation

```{r, warning = FALSE, message = FALSE}
#Libraries used:
library(readr)
library(ggplot2)
library(tidyverse)
```

```{r}
#Data loading:
data <- read_csv("Air_Quality.csv", show_col_types = FALSE)
glimpse(data)
```

```{r}
#Creation of subsets for every indicator:
data_1 <- data[data$Name == "Annual vehicle miles traveled",]
data_2 <- data[data$Name == "Annual vehicle miles traveled (trucks)",]
data_3_1 <- data[ (data$Name == "Asthma emergency departments visits due to Ozone") &
                   (data$Measure == "Estimated annual rate (age 18+)"),]
data_3_2 <- data[data$Name == "Asthma emergency departments visits due to Ozone" &
                  (data$Measure == "Estimated annual rate (under age 18)"),]
data_4 <- data[data$Name == "Boiler Emissions- Total NOx Emissions",]
data_5 <- data[data$Name == "Boiler Emissions- Total SO2 Emissions",]
data_6 <- data[data$Name == "Cardiovascular hospitalizations due to PM2.5 (age 40+)",]
data_7 <- data[data$Name == "Fine particles (PM 2.5)" & grepl("Annual Average", data$`Time Period`),]
data_8 <- data[data$Name == "Outdoor Air Toxics - Benzene",]
data_9 <- data[data$Name == "Ozone (O3)" & grepl("Summer 20", data$`Time Period`),]
data_10 <- data[data$Name == "Annual vehicle miles traveled (cars)",]
data_11_1 <- data[(data$Name == "Asthma emergency department visits due to PM2.5") &
                    (data$Measure == "Estimated annual rate (age 18+)"),]
data_11_2 <- data[data$Name == "Asthma emergency department visits due to PM2.5" &
                    (data$Measure == "Estimated annual rate (under age 18)"),]
data_12_1 <- data[data$Name == "Asthma hospitalizations due to Ozone" &
                  (data$Measure == "Estimated annual rate (age 18+)"),]
data_12_2 <- data[data$Name == "Asthma hospitalizations due to Ozone" &
                  (data$Measure == "Estimated annual rate (under age 18)"),]
data_13 <- data[data$Name == "Boiler Emissions- Total PM2.5 Emissions",]
data_14 <- data[data$Name == "Cardiac and respiratory deaths due to Ozone",]
data_15 <- data[data$Name == "Deaths due to PM2.5",]
data_16 <- data[data$Name == "Nitrogen dioxide (NO2)" & grepl("Annual Average", data$`Time Period`),]
data_17 <- data[data$Name == "Outdoor Air Toxics - Formaldehyde",]
data_18 <- data[data$Name == "Respiratory hospitalizations due to PM2.5 (age 20+)",]
```

```{r}
#Calculation of average values of indicators for every time period:
averaged_data_1 <- aggregate(data_1$`Data Value`, by = list(data_1$`Time Period`), FUN = mean)
averaged_data_2 <- aggregate(data_2$`Data Value`, by = list(data_2$`Time Period`), FUN = mean)
averaged_data_3_1 <- aggregate(data_3_1$`Data Value`, by = list(data_3_1$`Time Period`), FUN = mean)
averaged_data_3_2 <- aggregate(data_3_2$`Data Value`, by = list(data_3_2$`Time Period`), FUN = mean)
averaged_data_4 <- aggregate(data_4$`Data Value`, by = list(data_4$`Time Period`), FUN = mean)
averaged_data_5 <- aggregate(data_5$`Data Value`, by = list(data_5$`Time Period`), FUN = mean)
averaged_data_6 <- aggregate(data_6$`Data Value`, by = list(data_6$`Time Period`), FUN = mean)
averaged_data_7 <- aggregate(data_7$`Data Value`, by = list(data_7$`Time Period`), FUN = mean)
averaged_data_8 <- aggregate(data_8$`Data Value`, by = list(data_8$`Time Period`), FUN = mean)
averaged_data_9 <- aggregate(data_9$`Data Value`, by = list(data_9$`Time Period`), FUN = mean)
averaged_data_10 <- aggregate(data_10$`Data Value`, by = list(data_10$`Time Period`), FUN = mean)
averaged_data_11_1 <- aggregate(data_11_1$`Data Value`, by = list(data_11_1$`Time Period`), FUN = mean)
averaged_data_11_2 <- aggregate(data_11_2$`Data Value`, by = list(data_11_2$`Time Period`), FUN = mean)
averaged_data_12_1 <- aggregate(data_12_1$`Data Value`, by = list(data_12_1$`Time Period`), FUN = mean)
averaged_data_12_2 <- aggregate(data_12_2$`Data Value`, by = list(data_12_2$`Time Period`), FUN = mean)
averaged_data_13 <- aggregate(data_13$`Data Value`, by = list(data_13$`Time Period`), FUN = mean)
averaged_data_14 <- aggregate(data_14$`Data Value`, by = list(data_14$`Time Period`), FUN = mean)
averaged_data_15 <- aggregate(data_15$`Data Value`, by = list(data_15$`Time Period`), FUN = mean)
averaged_data_16 <- aggregate(data_16$`Data Value`, by = list(data_16$`Time Period`), FUN = mean)
averaged_data_17 <- aggregate(data_17$`Data Value`, by = list(data_17$`Time Period`), FUN = mean)
averaged_data_18 <- aggregate(data_18$`Data Value`, by = list(data_18$`Time Period`), FUN = mean)
```

```{r}
#Renaming columns for averaged dataframes:
colnames(averaged_data_1) <- c("Period", "Value")
colnames(averaged_data_2) <- c("Period", "Value")
colnames(averaged_data_3_1) <- c("Period", "Value")
colnames(averaged_data_3_2) <- c("Period", "Value")
colnames(averaged_data_4) <- c("Period", "Value")
colnames(averaged_data_5) <- c("Period", "Value")
colnames(averaged_data_6) <- c("Period", "Value")
colnames(averaged_data_7) <- c("Period", "Value")
colnames(averaged_data_8) <- c("Period", "Value")
colnames(averaged_data_9) <- c("Period", "Value")
colnames(averaged_data_10) <- c("Period", "Value")
colnames(averaged_data_11_1) <- c("Period", "Value")
colnames(averaged_data_11_2) <- c("Period", "Value")
colnames(averaged_data_12_1) <- c("Period", "Value")
colnames(averaged_data_12_2) <- c("Period", "Value")
colnames(averaged_data_13) <- c("Period", "Value")
colnames(averaged_data_14) <- c("Period", "Value")
colnames(averaged_data_15) <- c("Period", "Value")
colnames(averaged_data_16) <- c("Period", "Value")
colnames(averaged_data_17) <- c("Period", "Value")
colnames(averaged_data_18) <- c("Period", "Value")
```

## 1. Annual vehicle miles traveled
```{r}
#Table with values of indicator:
averaged_data_1

#Linear plot for indicator by time:
ggplot(averaged_data_1, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average annual vehicle miles traveled",
       x = "Time period",
       y = "Million miles per square mile")
```

Average annual vehicle miles traveled grew from 67.17757 million miles per
square mile	in 2005 to 79.96636 million miles per square mile in 2019. It tells
us about potential decrease of air quality.

## 2. Annual vehicle miles traveled (trucks)
```{r}
#Table with values of indicator:
averaged_data_2

#Linear plot for indicator by time:
ggplot(averaged_data_2, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average annual vehicle miles traveled (trucks)",
       x = "Time period",
       y = "Million miles per square mile")
```

Average annual vehicle miles traveled by trucks grew from 3.767290 million miles
per square mile	in 2005 to 4.477570 million miles per square mile in 2019. It
tells us about potential decrease of air quality.

## 3.1. Asthma emergency departments visits due to Ozone, age 18+
```{r}
#Table with values of indicator:
averaged_data_3_1

#Linear plot for indicator by time:
ggplot(averaged_data_3_1, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average asthma emergency departments visits due to Ozone, age 18+",
       x = "Time period",
       y = "Estimated annual rate per 100,000 adults")
```

Average asthma emergency departments visits due to Ozone fell from 48.41458 per
100,000 adults in 2005-2007 to 41.33333	per 100,000 adults in 2017-2019. It
tells us about potential increase of air quality.

## 3.2. Asthma emergency departments visits due to Ozone, under age 18
```{r}
#Table with values of indicator:
averaged_data_3_2

#Linear plot for indicator by time:
ggplot(averaged_data_3_2, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average asthma emergency departments visits due to Ozone, under age 18",
       x = "Time period",
       y = "Estimated annual rate per 100,000 children")
```

Average asthma emergency departments visits due to Ozone fell from 86.98958 per
100,000 children in 2005-2007 to 79.88679	per 100,000 children in 2017-2019. It
tells us about potential increase of air quality.

## 4. Boiler Emissions - Total NOx Emissions
```{r}
#Table with values of indicator:
averaged_data_4

#Linear plot for indicator by time:
ggplot(averaged_data_4, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average Boiler Emissions - Total NOx Emissions",
       x = "Time period",
       y = "Number per km2")
```

Average Boiler Emissions - Total NOx Emissions fell from 56.30000 per
km2 in 2013 to 51.28333 per km2 in 2015. It
tells us about potential increase of air quality.

## 5. Boiler Emissions- Total SO2 Emissions
```{r}
#Table with values of indicator:
averaged_data_5

#Linear plot for indicator by time:
ggplot(averaged_data_5, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average Boiler Emissions - Total SO2 Emissions",
       x = "Time period",
       y = "Number per km2")
```

Average Boiler Emissions - Total SO2 Emissions fell from 14.231250 per
km2 in 2013 to 7.752083 per km2 in 2015. It
tells us about potential increase of air quality.

## 6. Cardiovascular hospitalizations due to PM2.5, age 40+
```{r}
#Table with values of indicator:
averaged_data_6

#Linear plot for indicator by time:
ggplot(averaged_data_6, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average Cardiovascular hospitalizations due to PM2.5, age 40+",
       x = "Time period",
       y = "Estimated annual rate per 100,000 adults")
```

Average cardiovascular hospitalizations due to PM2.5 for people older than 40 years
fell from 26.150000 per 100,000 adults in 2005-2007 to 9.645833	per 100,000
adults in 2017-2019. It tells us about potential increase of air quality.

## 7. Fine particles (PM 2.5)
```{r}
#Table with values of indicator:
averaged_data_7$Period <- gsub("Annual Average ", "", averaged_data_7$Period)
averaged_data_7

#Linear plot for indicator by time:
ggplot(averaged_data_7, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average Fine particles (PM 2.5)",
       x = "Time period",
       y = "mcg/m3")
```

Average Fine particles (PM 2.5) fell from 10.986525 mcg/m3 in 2009 to 6.067376
mcg/m3 in 2022. It tells us about potential increase of air quality.

## 8. Outdoor Air Toxics - Benzene
```{r}
#Table with values of indicator:
averaged_data_8

#Linear plot for indicator by time:
ggplot(averaged_data_8, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average Outdoor Air Toxics - Benzene",
       x = "Time period",
       y = "Annual average concentration µg/m3")
```

Average Outdoor Air Toxics - Benzene fell from 2.910417 µg/m3 in 2005 to 0.925000
µg/m3 in 2014. It tells us about potential increase of air quality.

## 9. Ozone (O3)
```{r}
#Table with values of indicator:
averaged_data_9$Period <- gsub("Summer ", "", averaged_data_9$Period)
averaged_data_9

#Linear plot for indicator by time:
ggplot(averaged_data_9, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average Ozone (O3)",
       x = "Time period",
       y = "ppb")
```

Average Ozone (O3) grew from 24.76879 ppb in 2009 to 33.15603
ppb in 2022. It tells us about potential decrease of air quality.

## 10. Annual vehicle miles traveled (cars)
```{r}
#Table with values of indicator:
averaged_data_10

#Linear plot for indicator by time:
ggplot(averaged_data_10, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average Annual vehicle miles traveled (cars)",
       x = "Time period",
       y = "Million miles per square mile")
```

Average annual vehicle miles traveled by cars grew from 62.80374 million miles
per square mile	in 2005 to 72.91121 million miles per square mile in 2019. It
tells us about potential decrease of air quality.

## 11.1. Asthma emergency department visits due to PM2.5, age 18+
```{r}
#Table with values of indicator:
averaged_data_11_1

#Linear plot for indicator by time:
ggplot(averaged_data_11_1, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average asthma emergency department visits due to PM2.5, age 18+",
       x = "Time period",
       y = "Estimated annual rate per 100,000 adults")
```

Average asthma emergency departments visits due to PM2.5 fell from 61.17917 per
100,000 adults in 2005-2007 to 28.75000	per 100,000 adults in 2017-2019. It
tells us about potential increase of air quality.

## 11.2. Asthma emergency department visits due to PM2.5, under age 18
```{r}
#Table with values of indicator:
averaged_data_11_2

#Linear plot for indicator by time:
ggplot(averaged_data_11_2, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average asthma emergency department visits due to PM2.5, under age 18",
       x = "Time period",
       y = "Estimated annual rate per 100,000 children")
```

Average asthma emergency departments visits due to PM2.5 fell from 119.68542 per
100,000 children in 2005-2007 to 58.41667	per 100,000 children in 2017-2019. It
tells us about potential increase of air quality.

## 12.1. Asthma hospitalizations due to Ozone, age 18+
```{r}
#Table with values of indicator:
averaged_data_12_1

#Linear plot for indicator by time:
ggplot(averaged_data_12_1, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average asthma hospitalizations due to Ozone, age 18+",
       x = "Time period",
       y = "Estimated annual rate per 100,000 adults")
```

Average asthma hospitalizations due to Ozone fell from 7.514583 per 100,000
adults in 2005-2007 to 2.729167	per 100,000
adults in 2017-2019. It tells us about potential increase of air quality.

## 12.2 Asthma hospitalizations due to Ozone, under age 18
```{r}
#Table with values of indicator:
averaged_data_12_2

#Linear plot for indicator by time:
ggplot(averaged_data_12_2, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average asthma hospitalizations due to Ozone, under age 18",
       x = "Time period",
       y = "Estimated annual rate per 100,000 children")
```

Average asthma hospitalizations due to Ozone fell from 19.53333 per 100,000
children in 2005-2007 to 14.50000	per 100,000
children in 2017-2019. It tells us about potential increase of air quality.

## 13. Boiler Emissions - Total PM2.5 Emissions
```{r}
#Table with values of indicator:
averaged_data_13

#Linear plot for indicator by time:
ggplot(averaged_data_13, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average Boiler Emissions - Total PM2.5 Emissions",
       x = "Time period",
       y = "Number per km2")
```

Average Boiler Emissions - Total PM2.5 Emissions fell from 1.660417 per
km2 in 2013 to 1.087500 per km2 in 2015. It
tells us about potential increase of air quality.

## 14. Cardiac and respiratory deaths due to Ozone
```{r}
#Table with values of indicator:
averaged_data_14

#Linear plot for indicator by time:
ggplot(averaged_data_14, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average cardiac and respiratory deaths due to Ozone",
       x = "Time period",
       y = "Estimated annual rate per 100,000")
```

Average cardiac and respiratory deaths due to Ozone fell from 5.085417 per
100,000 in 2005-2007 to 4.854167 per 100,000 in 2017-2019. It
tells us about potential increase of air quality.

## 15. Deaths due to PM2.5
```{r}
#Table with values of indicator:
averaged_data_15

#Linear plot for indicator by time:
ggplot(averaged_data_15, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average deaths due to PM2.5, age 30+",
       x = "Time period",
       y = "Estimated annual rate per 100,000 adults")
```

Average deaths due to PM2.5 for people older than 30 years fell from 66.90833 per
100,000 adults in 2005-2007 to 33.91667 per 100,000 adults in 2017-2019. It
tells us about potential increase of air quality.

## 16. Nitrogen dioxide (NO2)
```{r}
#Table with values of indicator:
averaged_data_16$Period <- gsub("Annual Average ", "", averaged_data_16$Period)
averaged_data_16

#Linear plot for indicator by time:
ggplot(averaged_data_16, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average Nitrogen dioxide (NO2)",
       x = "Time period",
       y = "Mean ppb")
```

Average Nitrogen dioxide (NO2) fell from 25.75390 ppb in 2009 to 16.05461
ppb in 2022. It tells us about potential increase of air quality.

## 17. Outdoor Air Toxics - Formaldehyde
```{r}
#Table with values of indicator:
averaged_data_17

#Linear plot for indicator by time:
ggplot(averaged_data_17, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average Outdoor Air Toxics - Formaldehyde",
       x = "Time period",
       y = "Annual average concentration µg/m3")
```

Average Outdoor Air Toxics - Formaldehyde fell from 3.195833 µg/m3 in 2005 to
1.537500 µg/m3 in 2014. It tells us about potential increase of air quality.

## 18. Respiratory hospitalizations due to PM2.5, age 20+
```{r}
#Table with values of indicator:
averaged_data_18

#Linear plot for indicator by time:
ggplot(averaged_data_18, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average respiratory hospitalizations due to PM2.5, age 20+",
       x = "Time period",
       y = "Estimated annual rate per 100,000 adults")
```

Average respiratory hospitalizations due to PM2.5 for people older than 20 years
fell from 20.72917 per 100,000 adults in 2005-2007 to 7.43750	per 100,000
adults in 2017-2019. It tells us about potential increase of air quality.

## Conclusion

Almost all indicators of air quality in New York city were improved during
the analyzed period. Some deterioration is observed in miles traveled by
cars. However, increase in fuel efficiency of vehicles allowed to avoid
growth of emissions. Some increase is also observed in average Ozone (O3).
However, in general, air quality grew.






