---
title: "New York City Air Quality"
author: "Minh Tran"
date: "2024-11-26"
output: 
     slidy_presentation: default
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Introduction 
- New York City (NYC) Open Data, https://catalog.data.gov/dataset/air-quality
- Air Pollution - an ongoing, controversial and increasing problem.
- The indicators: Road Traffic, Ozone, PM 2.5, and Boiler Emissions
- Challenges ? Data Inconsistency
- Any interesting thing ? Harmful heath effects such as asthma, cardiac and respiratory deaths are included. 

---
# Outline {.bigger} 

1. Data Preparation
2. Road Traffic
3. Ozone
4. PM 2.5 (particles less than 2.5 micrometers in diameter)
5. Health Effects of Ozone and PM 2.5

# Data Preparation

```{r, echo=TRUE}
library(readr)
library(ggplot2)
library(tidyverse)
# For NYC map
library(nycgeo)
library(sf)
```

```{r}
#Data loading
data <- read_csv("Air_Quality.csv", show_col_types = FALSE)
glimpse(data)

```

```{r, echo = F}
#Creation of subsets for every indicator:
data_1 <- data[data$Name == "Annual vehicle miles traveled",]
data_2 <- data[data$Name == "Annual vehicle miles traveled (trucks)",]
data_3_1 <- data[ (data$Name == "Asthma emergency departments visits due to Ozone") &
                   (data$Measure == "Estimated annual rate (age 18+)"),]
data_3_2 <- data[data$Name == "Asthma emergency departments visits due to Ozone" &
                  (data$Measure == "Estimated annual rate (under age 18)"),]
data_4 <- data[data$Name == "Boiler Emissions- Total NOx Emissions",]
data_5 <- data[data$Name == "Boiler Emissions- Total SO2 Emissions",]
data_6 <- data[data$Name == "Cardiovascular hospitalizations due to PM2.5 (age 40+)",]
data_7 <- data[data$Name == "Fine particles (PM 2.5)" & grepl("Annual Average", data$`Time Period`),]
data_8 <- data[data$Name == "Outdoor Air Toxics - Benzene",]
data_9 <- data[data$Name == "Ozone (O3)" & grepl("Summer 20", data$`Time Period`),]
data_10 <- data[data$Name == "Annual vehicle miles traveled (cars)",]
data_11_1 <- data[(data$Name == "Asthma emergency department visits due to PM2.5") &
                    (data$Measure == "Estimated annual rate (age 18+)"),]
data_11_2 <- data[data$Name == "Asthma emergency department visits due to PM2.5" &
                    (data$Measure == "Estimated annual rate (under age 18)"),]
data_12_1 <- data[data$Name == "Asthma hospitalizations due to Ozone" &
                  (data$Measure == "Estimated annual rate (age 18+)"),]
data_12_2 <- data[data$Name == "Asthma hospitalizations due to Ozone" &
                  (data$Measure == "Estimated annual rate (under age 18)"),]
data_13 <- data[data$Name == "Boiler Emissions- Total PM2.5 Emissions",]
data_14 <- data[data$Name == "Cardiac and respiratory deaths due to Ozone",]
data_15 <- data[data$Name == "Deaths due to PM2.5",]
data_16 <- data[data$Name == "Nitrogen dioxide (NO2)" & grepl("Annual Average", data$`Time Period`),]
data_17 <- data[data$Name == "Outdoor Air Toxics - Formaldehyde",]
data_18 <- data[data$Name == "Respiratory hospitalizations due to PM2.5 (age 20+)",]
```

```{r, echo = F}
#Calculation of average values of indicators for every time period:
averaged_data_1 <- aggregate(data_1$`Data Value`, by = list(data_1$`Time Period`), FUN = mean)
averaged_data_2 <- aggregate(data_2$`Data Value`, by = list(data_2$`Time Period`), FUN = mean)
averaged_data_3_1 <- aggregate(data_3_1$`Data Value`, by = list(data_3_1$`Time Period`), FUN = mean)
averaged_data_3_2 <- aggregate(data_3_2$`Data Value`, by = list(data_3_2$`Time Period`), FUN = mean)
averaged_data_4 <- aggregate(data_4$`Data Value`, by = list(data_4$`Time Period`), FUN = mean)
averaged_data_5 <- aggregate(data_5$`Data Value`, by = list(data_5$`Time Period`), FUN = mean)
averaged_data_6 <- aggregate(data_6$`Data Value`, by = list(data_6$`Time Period`), FUN = mean)
averaged_data_7 <- aggregate(data_7$`Data Value`, by = list(data_7$`Time Period`), FUN = mean)
averaged_data_8 <- aggregate(data_8$`Data Value`, by = list(data_8$`Time Period`), FUN = mean)
averaged_data_9 <- aggregate(data_9$`Data Value`, by = list(data_9$`Time Period`), FUN = mean)
averaged_data_10 <- aggregate(data_10$`Data Value`, by = list(data_10$`Time Period`), FUN = mean)
averaged_data_11_1 <- aggregate(data_11_1$`Data Value`, by = list(data_11_1$`Time Period`), FUN = mean)
averaged_data_11_2 <- aggregate(data_11_2$`Data Value`, by = list(data_11_2$`Time Period`), FUN = mean)
averaged_data_12_1 <- aggregate(data_12_1$`Data Value`, by = list(data_12_1$`Time Period`), FUN = mean)
averaged_data_12_2 <- aggregate(data_12_2$`Data Value`, by = list(data_12_2$`Time Period`), FUN = mean)
averaged_data_13 <- aggregate(data_13$`Data Value`, by = list(data_13$`Time Period`), FUN = mean)
averaged_data_14 <- aggregate(data_14$`Data Value`, by = list(data_14$`Time Period`), FUN = mean)
averaged_data_15 <- aggregate(data_15$`Data Value`, by = list(data_15$`Time Period`), FUN = mean)
averaged_data_16 <- aggregate(data_16$`Data Value`, by = list(data_16$`Time Period`), FUN = mean)
averaged_data_17 <- aggregate(data_17$`Data Value`, by = list(data_17$`Time Period`), FUN = mean)
averaged_data_18 <- aggregate(data_18$`Data Value`, by = list(data_18$`Time Period`), FUN = mean)
```

```{r, echo = F}
#Renaming columns for averaged dataframes:
colnames(averaged_data_1) <- c("Period", "Value")
colnames(averaged_data_2) <- c("Period", "Value")
colnames(averaged_data_3_1) <- c("Period", "Value")
colnames(averaged_data_3_2) <- c("Period", "Value")
colnames(averaged_data_4) <- c("Period", "Value")
colnames(averaged_data_5) <- c("Period", "Value")
colnames(averaged_data_6) <- c("Period", "Value")
colnames(averaged_data_7) <- c("Period", "Value")
colnames(averaged_data_8) <- c("Period", "Value")
colnames(averaged_data_9) <- c("Period", "Value")
colnames(averaged_data_10) <- c("Period", "Value")
colnames(averaged_data_11_1) <- c("Period", "Value")
colnames(averaged_data_11_2) <- c("Period", "Value")
colnames(averaged_data_12_1) <- c("Period", "Value")
colnames(averaged_data_12_2) <- c("Period", "Value")
colnames(averaged_data_13) <- c("Period", "Value")
colnames(averaged_data_14) <- c("Period", "Value")
colnames(averaged_data_15) <- c("Period", "Value")
colnames(averaged_data_16) <- c("Period", "Value")
colnames(averaged_data_17) <- c("Period", "Value")
colnames(averaged_data_18) <- c("Period", "Value")
```

# Road Traffic
```{r echo = F, fig.height= 4, message=FALSE}
library(tidyverse)

traffic_plot <- ggplot(data = averaged_data_10, aes(x = Period, y = Value, group = 1,fill = "cars")) +
  geom_col(width = 0.5) +
  geom_col(data = averaged_data_2, aes(x = Period, y = Value, group = 1,fill = "trucks"), width = 0.5) +
  
  
  labs(title = "Average annual vehicle miles traveled",
       x = "Time period",
       y = "Million miles per square mile", fill = "Vehicles")

traffic_plot

```

- Trucks: 3.7 to 4.4   
- Cars: 62.8 to 72.9




```{r fig.width= 10, echo = F, fig.height= 5}
library(cowplot)
#Remove Annual Average string in the data for NO2 plot  
averaged_data_16$Period <- gsub("Annual Average ", "", averaged_data_16$Period)

# These two substances have the same unit of measurement, and time of observation so I put them into one graph. 
Benzene_Formaldehyde <- ggplot(averaged_data_8, aes(x = Period, y = Value, group = 1,color = "Benzene")) +
  geom_line(size = 1) + 
  geom_point(data = averaged_data_8, aes(x = Period, y = Value, group = 1,color = "Benzene"),size = 3) +
  
  geom_line(data = averaged_data_17, aes(x = Period, y = Value, group = 1,color = "Formaldehyde"),size = 1) +
  geom_point(data = averaged_data_17, aes(x = Period, y = Value, group = 1,color = "Formaldehyde"),size = 3) +

  labs(title = "Concentration in Air ", subtitle = "Formaldehyde and Benzene",
       x = "Time period",
       y = "Annual average concentration ug/m3", color = "Substances") +
   theme(legend.position = c(0.8, 0.8))

# NO2 graph on the right
NO2 <- ggplot(averaged_data_16, aes(x = Period, y = Value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Average Nitrogen dioxide (NO2)",
       x = "Time period",
       y = "Mean ppb") 

# Combine two graphs into one horizontal arrangement
plot_grid(Benzene_Formaldehyde, NO2, labels = NA)

```
- Road traffic is not a significant in air pollution at NYC

# Ozone
- Ozone concentration around NYC in 2009 and 2002
```{r,fig.width= 9.6, echo=FALSE}
library(nycgeo)
library(sf)
library(cowplot)

ozone3_2009 <- data %>% 
  filter(Name == "Ozone (O3)",`Time Period` == "Summer 2009")
ozone3_2022 <- data %>% 
  filter(Name == "Ozone (O3)",`Time Period` == "Summer 2022")

# Change data type for `Geo Join ID` to char before joining 
ozone3_2009 <- ozone3_2009 %>% 
  mutate(`Geo Join ID`= as.character(`Geo Join ID`))
ozone3_2022 <- ozone3_2022 %>% 
  mutate(`Geo Join ID`= as.character(`Geo Join ID`))

# Graphing 2009 
ozone_2009_graph <- nyc_boundaries(geography = "cd") %>% 
  inner_join(ozone3_2009, by = c("borough_cd_id" = "Geo Join ID")) %>% 
  ggplot() +
  geom_sf(aes(fill = `Data Value`)) +
  scale_fill_viridis_c(name = "ppb", option = "inferno", direction = -1) +
  theme_void() +
  theme(panel.grid = element_line(color = "transparent"),legend.position = c(0.1, 0.7)) +
  labs(title = "2009")

# Graphing 2022
ozone_2022_graph <- nyc_boundaries(geography = "cd") %>% 
  inner_join(ozone3_2022, by = c("borough_cd_id" = "Geo Join ID")) %>% 
  ggplot() +
  geom_sf(aes(fill = `Data Value`)) +
  scale_fill_viridis_c(name = "ppb", option = "inferno", direction = -1) +
  theme_void() +
  theme(panel.grid = element_line(color = "transparent"),legend.position = c(0.1, 0.7)) +
  labs(title = "2022")

# Combine 2009 and 2022
plot_grid(ozone_2009_graph, ozone_2022_graph, labels = NA)

```


- Which area had the lowest Ozone concentration in 2009 ? How was that amount in 2022 ?
```{r, echo=FALSE, fig.width= 9.95}
k9 <- data %>%
  filter(Name == "Ozone (O3)",`Time Period` == "Summer 2009") %>% 
  slice_min(`Data Value`, n =1) %>% 
  select(`Geo Join ID`, `Geo Place Name`, ppb_amout_2009 = `Data Value`)

k9
```

```{r, echo=FALSE, fig.width= 9.95}
k22 <- data %>%
  filter(Name == "Ozone (O3)",`Geo Place Name` == "Midtown (CD5)",`Time Period` == "Summer 2022") %>%
  select(`Geo Join ID`, ppb_amout_2022 = `Data Value`)


compare_9_22 <- inner_join(k9,k22, by = "Geo Join ID")


with_change <- compare_9_22 %>% 
  mutate(Percentage_Change = ppb_amout_2022/ppb_amout_2009*100)

knitr::kable(with_change, caption = "Change of Ozone Concentration for Midtown (CD5) from 2009 to 2022")

```

# PM 2.5
- Average amount of PM 2.5 from 2009 to 2022. 
```{r, echo=FALSE, fig.height= 5}
averaged_data_7$Period <- gsub("Annual Average ", "", averaged_data_7$Period)
averaged_data_7

# Plot for PM 2.5
ggplot(averaged_data_7, aes(x = Period, y = Value, group = 1)) +
  geom_col() +
  labs(title = "Average Fine particles (PM 2.5)",
       x = "Year",
       y = "mcg/m3")


```

- Dangerous level of PM 2.5: above 35 micrograms per cubic meter (µg/m3) or mcg/m3 (U.S. Environmental Protection Agency (EPA))

# Ozone and PM 2.5, Which one is more harmful to humans?
- Asthma emergency department visits
```{r, echo=FALSE, fig.width= 9.6, fig.height= 5}
library(cowplot)
# Above 18 Asthma emergency department visits data
add_colum_1 <- averaged_data_3_1 %>%   # by Ozone, above_18
  mutate(Indicator = c("Ozone","Ozone","Ozone","Ozone","Ozone"))




add_colum_2 <- averaged_data_11_1 %>% # by PM25, above_18  
  mutate(Indicator = c("PM25","PM25","PM25","PM25","PM25"))


pm_ozone_above_18 <- full_join(add_colum_1,add_colum_2, by = NULL)


# Above 18 Asthma emergency department visits graph
above_18 <- ggplot() +
  geom_col(data = pm_ozone_above_18, aes(x = Period, y = Value, fill = Indicator),
           position = "dodge") +
  scale_fill_manual(values= c("#F8B195", "#355C7D"))+
  labs(title = "Adults ",
       x = "Time period",
       y = "Per 100,000 adults")+
  theme(legend.position = c(0.9,0.9))


# Below 18 Data
add_colum_3 <- averaged_data_3_2 %>%   # Asthma by Ozone, below 18
  mutate(Indicator = c("Ozone","Ozone","Ozone","Ozone","Ozone"))




add_colum_4 <- averaged_data_11_2 %>% # Asthma by PM25, above_18  
  mutate(Indicator = c("PM25","PM25","PM25","PM25","PM25"))


pm_ozone_below_18 <- full_join(add_colum_3,add_colum_4, by = NULL)


# Below 18 graph
below_18 <- ggplot() +
  geom_col(data = pm_ozone_below_18, aes(x = Period, y = Value, fill = Indicator),
           position = "dodge") +
  scale_fill_manual(values= c("#F8B195", "#355C7D"))+
  labs(title = "Children ",
       x = "Time period",
       y = "Per 100,000 children")+
  theme(legend.position = c(0.9,0.9))

# Combine
plot_grid(below_18, above_18, labels = NA)


```

- Deaths due to Ozone and PM 2.5
```{r, echo=FALSE, fig.width= 9.6, fig.height= 5}


# Graph
ggplot(data_14, aes(`Time Period`, `Data Value`)) +
  geom_point() +
  geom_smooth(data = data_14, aes(`Time Period`, `Data Value`,group=-1, color = "Ozone")) +
  geom_point(data = data_15, aes(`Time Period`, `Data Value`)) +
  
  geom_smooth(data = data_15, aes(`Time Period`, `Data Value`,group=-1, color = "PM 2.5")) +
  labs(title = "Deaths rates due to Ozone and PM2.5",
       x = "Time period",
       y = "Per 100,000 people", color = "") +
  theme(legend.position = c(0.9,0.95))
```